---
title: "TraitAM Functional Analysis Report"
author: "Generated by Seqera AI TraitAM Pipeline"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: hide
    df_print: paged
params:
  sample_id: "sample1"
  results_dir: "results"
  traitam_dir: "data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.width = 10,
    fig.height = 7
)

library(tidyverse)
library(knitr)
library(DT)
library(ggpubr)
library(patchwork)
```

```{r load_data}
# Load all analysis results
sample_id <- params$sample_id
results_dir <- params$results_dir

# Matched taxa
matched_taxa <- read_tsv(
    file.path(results_dir, "01_matched_taxa", 
              paste0(sample_id, "_matched_species.tsv")),
    show_col_types = FALSE
)

# Trait data
traits_data <- read_tsv(
    file.path(results_dir, "02_trait_data", 
              paste0(sample_id, "_traits_abundance.tsv")),
    show_col_types = FALSE
)

# FD metrics
fd_metrics <- read_tsv(
    file.path(results_dir, "03_functional_diversity", 
              paste0(sample_id, "_FD_metrics.tsv")),
    show_col_types = FALSE
)

# CWM
cwm_data <- read_tsv(
    file.path(results_dir, "03_functional_diversity", 
              paste0(sample_id, "_CWM.tsv")),
    show_col_types = FALSE
)

# Phylogenetic signal (if exists)
phylo_signal_file <- file.path(results_dir, "04_phylogenetic_analysis", 
                                paste0(sample_id, "_phylo_signal.tsv"))
if (file.exists(phylo_signal_file)) {
    phylo_signal <- read_tsv(phylo_signal_file, show_col_types = FALSE)
} else {
    phylo_signal <- NULL
}

# Phylogenetic diversity (if exists)
phylo_div_file <- file.path(results_dir, "04_phylogenetic_analysis", 
                             paste0(sample_id, "_phylo_diversity.tsv"))
if (file.exists(phylo_div_file)) {
    phylo_diversity <- read_tsv(phylo_div_file, show_col_types = FALSE)
} else {
    phylo_diversity <- NULL
}
```

---

# Executive Summary {.tabset}

## Overview

This report presents a comprehensive functional trait analysis of Arbuscular Mycorrhizal Fungi (AMF) detected in sample **`r sample_id`**, integrating data from the **TraitAM** global spore trait database.

### Key Findings:

- **Species Detected:** `r n_distinct(matched_taxa$species_name)` species matched to TraitAM
- **Functional Richness (FRic):** `r round(fd_metrics$FRic, 3)`
- **Functional Evenness (FEve):** `r round(fd_metrics$FEve, 3)`
- **Functional Divergence (FDiv):** `r round(fd_metrics$FDiv, 3)`
- **Phylogenetic Diversity (PD):** `r if(!is.null(phylo_diversity)) round(phylo_diversity$PD, 2) else "N/A"`

### Analysis Date:

Generated on `r Sys.Date()` using TraitAM database (v2024).

---

## Match Statistics

```{r match_stats}
match_summary <- matched_taxa %>%
    count(match_type, name = "Count") %>%
    mutate(
        Percentage = round(100 * Count / sum(Count), 1),
        Match_Type = case_when(
            match_type == "exact_species" ~ "Exact species match",
            grepl("fuzzy", match_type) ~ "Fuzzy species match",
            match_type == "genus_level" ~ "Genus-level match",
            TRUE ~ "Unmatched"
        )
    ) %>%
    select(Match_Type, Count, Percentage)

kable(match_summary, 
      caption = "Taxonomic Matching Summary",
      col.names = c("Match Type", "Count", "Percentage (%)"))
```

**Total OTUs analyzed:** `r nrow(matched_taxa)`  
**Successfully matched:** `r sum(!is.na(matched_taxa$match_type))` (`r round(100*sum(!is.na(matched_taxa$match_type))/nrow(matched_taxa), 1)`%)

---

# Taxonomic Composition {.tabset}

## Species Table

```{r species_table}
species_abundance <- traits_data %>%
    group_by(species_name, Family, Genus) %>%
    summarise(
        Total_Abundance = sum(abundance),
        Relative_Abundance = round(100 * Total_Abundance / sum(traits_data$abundance), 2),
        .groups = "drop"
    ) %>%
    arrange(desc(Total_Abundance))

datatable(
    species_abundance,
    options = list(pageLength = 15, scrollX = TRUE),
    caption = "Detected Species with Abundance Data",
    rownames = FALSE,
    colnames = c("Species", "Family", "Genus", "Abundance", "Rel. Abundance (%)")
)
```

## Family Composition

```{r family_composition, fig.height=6}
family_abundance <- traits_data %>%
    group_by(Family) %>%
    summarise(Total_Abundance = sum(abundance), .groups = "drop") %>%
    mutate(Percentage = 100 * Total_Abundance / sum(Total_Abundance)) %>%
    arrange(desc(Percentage))

ggplot(family_abundance, aes(x = reorder(Family, Percentage), y = Percentage, fill = Family)) +
    geom_col() +
    coord_flip() +
    scale_fill_viridis_d() +
    theme_bw() +
    labs(
        title = "Community Composition by Family",
        x = "Family",
        y = "Relative Abundance (%)"
    ) +
    theme(legend.position = "none")
```

---

# Functional Traits {.tabset}

## Trait Summary

```{r trait_summary}
trait_summary <- traits_data %>%
    select(Spore_vol_um3, Aspect_ratio, Color_score, 
           Wall_investment, Ornamentation_height_um) %>%
    summarise(across(everything(), list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
    ))) %>%
    pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
    separate(Statistic, into = c("Trait", "Stat"), sep = "_(?=[^_]+$)") %>%
    pivot_wider(names_from = Stat, values_from = Value)

kable(trait_summary, 
      digits = 2,
      caption = "Summary Statistics for Functional Traits")
```

## Community Weighted Means

```{r cwm_table}
cwm_display <- cwm_data %>%
    pivot_longer(-Sample, names_to = "Trait", values_to = "CWM") %>%
    mutate(
        Trait_Name = case_when(
            Trait == "Spore_vol_um3" ~ "Spore Volume (μm³)",
            Trait == "Aspect_ratio" ~ "Aspect Ratio",
            Trait == "Color_score" ~ "Color Score",
            Trait == "Wall_investment" ~ "Wall Investment",
            Trait == "Ornamentation_height_um" ~ "Ornamentation Height (μm)",
            TRUE ~ Trait
        ),
        CWM = round(CWM, 3)
    ) %>%
    select(Trait_Name, CWM)

kable(cwm_display,
      col.names = c("Trait", "Community Weighted Mean"),
      caption = "Abundance-Weighted Average Trait Values")
```

**Interpretation:** CWM values represent the dominant trait values in the community, weighted by species abundance.

## Trait Distributions

```{r trait_distributions, fig.height=8}
knitr::include_graphics(
    file.path(results_dir, "06_visualizations", 
              paste0(sample_id, "_trait_distribution.pdf"))
)
```

## Trait Correlations

```{r trait_correlations}
knitr::include_graphics(
    file.path(results_dir, "06_visualizations", 
              paste0(sample_id, "_trait_correlations.pdf"))
)
```

---

# Functional Diversity {.tabset}

## FD Metrics

```{r fd_metrics_table}
fd_display <- fd_metrics %>%
    select(FRic, FEve, FDiv, FDis, RaoQ) %>%
    pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
    mutate(
        Description = case_when(
            Metric == "FRic" ~ "Functional richness (volume of trait space)",
            Metric == "FEve" ~ "Functional evenness (regularity of spacing)",
            Metric == "FDiv" ~ "Functional divergence (degree of niche differentiation)",
            Metric == "FDis" ~ "Functional dispersion (mean distance to centroid)",
            Metric == "RaoQ" ~ "Rao's quadratic entropy (trait diversity)",
            TRUE ~ ""
        ),
        Value = round(Value, 4)
    )

kable(fd_display,
      col.names = c("Index", "Value", "Description"),
      caption = "Functional Diversity Metrics")
```

### Interpretation:

- **FRic** measures the total volume of functional trait space occupied by the community
- **FEve** indicates how evenly species are distributed in trait space
- **FDiv** quantifies how abundance is distributed at the extremes of trait space
- **FDis** represents the mean distance of species to the centroid in trait space
- **RaoQ** combines trait differences and species abundances

## PCA Analysis

```{r pca_biplot}
knitr::include_graphics(
    file.path(results_dir, "06_visualizations", 
              paste0(sample_id, "_PCA_biplot.pdf"))
)
```

## Functional Space

```{r functional_space}
func_space_file <- file.path(results_dir, "03_functional_diversity", 
                              paste0(sample_id, "_functional_space.pdf"))
if (file.exists(func_space_file)) {
    knitr::include_graphics(func_space_file)
}
```

---

# Phylogenetic Analysis {.tabset}

```{r phylo_check}
has_phylo <- !is.null(phylo_signal) && !is.null(phylo_diversity)
```

`r if(!has_phylo) "**Note:** Phylogenetic analysis was not performed (insufficient species matched to tree)."`

## Phylogenetic Signal {.tabset}

`r if(has_phylo) "### Signal Statistics"`

```{r phylo_signal_table}
if (!is.null(phylo_signal)) {
    phylo_display <- phylo_signal %>%
        mutate(
            Blomberg_K = round(Blomberg_K, 3),
            K_pvalue = round(K_pvalue, 4),
            Pagel_lambda = round(Pagel_lambda, 3),
            Lambda_pvalue = round(Lambda_pvalue, 4)
        )
    
    kable(phylo_display,
          caption = "Phylogenetic Signal of Functional Traits",
          col.names = c("Trait", "Blomberg's K", "K p-value", 
                       "Pagel's λ", "λ p-value", "Interpretation"))
}
```

`r if(has_phylo) "**Interpretation:**"`

`r if(has_phylo) "- **K < 0.5:** Weak phylogenetic signal (traits evolve independently)"`
`r if(has_phylo) "- **K ≈ 1:** Traits evolve under Brownian motion (expected under neutral evolution)"`
`r if(has_phylo) "- **K > 1:** Strong phylogenetic signal (trait conservatism, closely related species are similar)"`

## Phylogenetic Diversity

```{r phylo_diversity_table}
if (!is.null(phylo_diversity)) {
    pd_display <- phylo_diversity %>%
        mutate(
            PD = round(PD, 2),
            MPD = round(MPD, 3),
            MNTD = round(MNTD, 3)
        ) %>%
        select(PD, SR, MPD, MNTD)
    
    kable(pd_display,
          caption = "Phylogenetic Diversity Metrics",
          col.names = c("Faith's PD", "Species Richness", "MPD", "MNTD"))
}
```

`r if(has_phylo) "**Metrics:**"`

`r if(has_phylo) "- **PD (Faith's Phylogenetic Diversity):** Total branch length in phylogeny"`
`r if(has_phylo) "- **MPD (Mean Pairwise Distance):** Average phylogenetic distance between all species pairs"`
`r if(has_phylo) "- **MNTD (Mean Nearest Taxon Distance):** Average distance to closest relative"`

## Phylogenetic Tree

```{r phylo_tree}
tree_file <- file.path(results_dir, "04_phylogenetic_analysis", 
                       paste0(sample_id, "_tree_comparison.pdf"))
if (file.exists(tree_file)) {
    knitr::include_graphics(tree_file)
}
```

---

# Trait-Abundance Relationships {.tabset}

## Overview

Do abundant species exhibit specific trait values? This section explores correlations between species abundance and functional traits.

## Scatterplots

```{r abundance_traits}
knitr::include_graphics(
    file.path(results_dir, "06_visualizations", 
              paste0(sample_id, "_abundance_traits.pdf"))
)
```

## Statistical Tests

```{r abundance_correlations}
# Calculate correlations between log abundance and each trait
cor_results <- traits_data %>%
    distinct(species_name, .keep_all = TRUE) %>%
    select(abundance, Spore_vol_um3, Aspect_ratio, Color_score, 
           Wall_investment, Ornamentation_height_um) %>%
    mutate(log_abundance = log10(abundance + 1)) %>%
    select(-abundance) %>%
    summarise(across(-log_abundance, ~cor(.x, log_abundance, use = "complete.obs"))) %>%
    pivot_longer(everything(), names_to = "Trait", values_to = "Correlation") %>%
    mutate(
        Correlation = round(Correlation, 3),
        Interpretation = case_when(
            abs(Correlation) > 0.7 ~ "Strong",
            abs(Correlation) > 0.4 ~ "Moderate",
            abs(Correlation) > 0.2 ~ "Weak",
            TRUE ~ "Very weak"
        )
    )

kable(cor_results,
      caption = "Pearson Correlation: log(Abundance) vs Traits",
      col.names = c("Trait", "Correlation (r)", "Strength"))
```

---

# Data Tables {.tabset}

## Complete Trait Data

```{r complete_trait_table}
trait_table <- traits_data %>%
    select(species_name, Family, abundance, 
           Spore_vol_um3, Aspect_ratio, Color_score, 
           Wall_investment, Ornamentation_height_um) %>%
    mutate(across(where(is.numeric), ~round(.x, 2)))

datatable(
    trait_table,
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Complete Trait and Abundance Data",
    filter = "top"
)
```

## Matched Taxa

```{r matched_taxa_table}
datatable(
    matched_taxa %>% select(name, taxopath, LWR, species_name, match_type),
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "OTU to TraitAM Species Matching",
    filter = "top",
    colnames = c("OTU ID", "Taxonomy Path", "Confidence", "Matched Species", "Match Type")
)
```

---

# Methods

## Pipeline Overview

1. **Taxonomic Matching:** OTU taxonomy from GAPPA was matched to TraitAM species database using exact, fuzzy, and genus-level matching strategies.

2. **Trait Extraction:** Functional trait data (spore volume, aspect ratio, color, wall investment, ornamentation) extracted from TraitAM DataRecord_1.

3. **Abundance Integration:** OTU abundance data integrated with trait values.

4. **Functional Diversity:** Calculated using the `FD` package (Laliberté & Legendre 2010):
   - FRic, FEve, FDiv, FDis, RaoQ
   - Community weighted means (CWM)

5. **Phylogenetic Analysis:** Using `picante` package (Kembel et al. 2010):
   - Phylogenetic signal (Blomberg's K, Pagel's λ)
   - Phylogenetic diversity (Faith's PD, MPD, MNTD)

## Software

- **R** version `r paste(R.version$major, R.version$minor, sep = ".")`
- **Packages:** tidyverse, FD, vegan, ape, phytools, picante, ggtree

## TraitAM Database

**Version:** June 2024  
**Citation:** Chaudhary et al. (2024). TraitAM, a global spore trait and phylogeny database for arbuscular mycorrhizal fungi.  
**URL:** https://datadryad.org/dataset/doi:10.5061/dryad.6hdr7sr8z

---

# References

1. **TraitAM Database:**  
   Chaudhary, V. B., et al. (2024). TraitAM, a global spore trait and phylogeny database for arbuscular mycorrhizal fungi. *Manuscript submitted.*

2. **FD Package:**  
   Laliberté, E. & Legendre, P. (2010). A distance-based framework for measuring functional diversity from multiple traits. *Ecology* 91:299-305.

3. **picante Package:**  
   Kembel, S.W., et al. (2010). Picante: R tools for integrating phylogenies and ecology. *Bioinformatics* 26:1463-1464.

4. **vegan Package:**  
   Oksanen, J., et al. (2020). vegan: Community Ecology Package. R package version 2.5-7.

5. **ggtree:**  
   Yu, G., et al. (2017). ggtree: an R package for visualization and annotation of phylogenetic trees. *Methods in Ecology and Evolution* 8:28-36.

---

# Session Info

```{r session_info}
sessionInfo()
```

---

**Report generated:** `r Sys.time()`  
**Pipeline:** TraitAM Functional Analysis v1.0.0  
**Sample:** `r sample_id`

